Life in the canopy: Phylogenetic filtering and morphological canalization in epiphytic ferns

### 2018-03-23 manuscript.Rnw
Fixes to mean_trait_values
Move sourcing of mean_trait_values to top of analyses list, add gameto_source to mean_trait_values, fix caption in supporting information Table S5

<<run_analyses, results="hide">>=
# set working directory, run all analyses, save results objects
setwd(here::here())
source(file = "bin/microclimate.R")
source(file = "bin/trait_PCA.R")
source(file = "bin/cont_phylosig.R")
source(file = "bin/binary_phylosig.R")
source(file = "bin/sporo_PIC.R")
source(file = "bin/correlated_evo.R")
source(file = "bin/binaryPGLMM.R")
source(file = "bin/functional_diversity.R")
source(file = "bin/mean_trait_values.R") # this must be last to keep sla.grand.mean
save(list = ls(all = TRUE), file= "results.RData")
@

@@

\begin{itemize}
\item The appearance of angiosperm-dominated forests in the Cretaceous and early Cenozoic had a profound effect on terrestrial biota by creating novel ecological niches. The diversity of modern fern lineages are thought to be a response to angiosperm dominance, radiating both in the canopy and on the forest floor. Despite the significance of epiphytic growth for fern evolution, however, few studies have investigated traits related to epiphytism in ferns while accounting for phylogeny.
\item We investigated evolutionary patterns in traits related to epiphytic growth at the species and community level using a phylogeny including representatives of multiple epiphytic fern radiations on the island of Moorea, French Polynesia. 
\item We observed phylogenetic conservatism in multiple traits related to epiphytic growth in fern sporophytes and gametophytes. Epiphytes had significantly shorter stipes, smaller rhizomes, and a higher frequency of non-cordate gametophytes while accounting for phylogeny. Epiphytic communities were clustered both phylogenetically and functionally, consistent with the hypothesis that the canopy acts as a filter due to its low humidity. 
\item Our results provide an important phylogenetic perspective on the role of functional traits in ferns, and pave the way for more detailed studies examining how niche space is filled within terrestrial and epiphytic habitats.
\end{itemize}

@@

The sporophyte traits that differed most strongly between epiphytic and terrestrial taxa were related to overall size, including stipe and frond length, frond width, and rhizome diameter. Stipe length in particular differed strongly between epiphytes and terrestrial species (\fig\ref{fig:CWM}B, \tab\ref{table:PIC_sporo}). Stipe length may be a critical trait controlling whole leaf hydraulic conductance; since ferns have resistive stipes in general, species growing in water-limited habitats are expected to decrease stipe length to minimize resistance \citep{WatkinsJr.2010}. Our results support previous non-phylogenetically controlled studies that found a prevalence of short stipes in epiphytic ferns \citep{WatkinsJr.2010,Creese2011}. However, given that many of these size-related traits scale together in ferns \citep{Arcand2008,Creese2011}, careful experimental manipulations are needed to distinguish the role (or lack thereof) of each in epiphytic growth.

Surprisingly, SLA was only slightly lower in epiphytes relative to terrestrial species (\tab\ref{table:PIC_sporo}, \fig\ref{fig:CWM}B). SLA is an important part of the leaf economic spectrum, which posits that leaf traits including SLA, N and C content, and leaf life span vary along well-defined axes such that plants generally fall into one of two categories: fast-growing species with high SLA, low nutrient content, and short life spans vs. slow-growing species with low SLA, high nutrient content, and long life spans \citep{Wright2004a}. We hypothesized that epiphytes would have lower SLA, given their more extreme growing conditions. Indeed, the species with the lowest SLA values were epiphytes (\eg, \textit{Selliguea plantaginea} Brack., \SI{\Sexpr{as.numeric(get_result(filter(sla.grand.mean, species=="Selliguea_plantaginea"), "mean", signif=TRUE))}}{\square\meter\per\kilo\gram}; \textit{Ctenopterella blechnoides} (Grev.) Parris, \SI{\Sexpr{as.numeric(get_result(filter(sla.grand.mean, species=="Ctenopterella_blechnoides"), "mean", signif=TRUE))}}{\square\meter\per\kilo\gram}), but this relationship was not significant across the dataset (\tab\ref{table:PIC_sporo}, \fig\ref{fig:CWM}). The tradeoffs posited in the leaf economics spectrum have been most thoroughly documented in global studies including the widest possible range of taxa and environments, and do not always hold up in local studies \citep{Petter2016}. For example, \citet{Zhu2016} did not find a significant relationship between SLA and leaf life span in a survey of 16 ferns across a disturbance (light) gradient in China. It is possible that other small-scale effects (\eg, microclimate) obscure this pattern in local studies such as ours and that of \citet{Zhu2016}. This is supported by the strong, parallel relationship in CWM values of SLA with elevation in both epiphytes and terrestrial species (\fig\ref{fig:CWM}A).

@@

% Table S5 
\begin{table}[ht]
\caption{List of all ferns from Moorea, French Polynesia with observed trait values and growth habit. For continuous trait values, means $\pm$ \stdev shown. For growth habit, E = epiphytic, T = terrestrial. For frond dissection, 1 = simple, 2 = pinnatifid or pinnatisect, 3 = 1-pinnate, 4 = 1- pinnate-pinnatifid, 5 = 2-pinnate, 6 = 2-pinnate-pinnatifid, 7 = 3- pinnate, 8 = 3-pinnate-pinnatifid, 9 = \textgreater 3-pinnate-pinnatifid, 10 = binpinnatifid or tripinnatifid.}
\label{supp_table:traitvals}
\end{table}

>>>>

<<run_analyses, results="hide">>=
# set working directory, run all analyses, save results objects
setwd(here::here())
source(file = "bin/mean_trait_values.R")
source(file = "bin/microclimate.R")
source(file = "bin/trait_PCA.R")
source(file = "bin/cont_phylosig.R")
source(file = "bin/binary_phylosig.R")
source(file = "bin/sporo_PIC.R")
source(file = "bin/correlated_evo.R")
source(file = "bin/binaryPGLMM.R")
source(file = "bin/functional_diversity.R")
save(list = ls(all = TRUE), file= "results.RData")
@

@@

\begin{itemize}
\item The appearance of angiosperm-dominated forests in the Cretaceous and early Cenozoic had a profound effect on terrestrial biota by creating novel ecological niches. The diversity of modern fern lineages is thought to be a response to angiosperm dominance, radiating both in the canopy and on the forest floor. Despite the significance of epiphytic growth for fern evolution, however, few studies have investigated traits related to epiphytism in ferns while accounting for phylogeny.
\item We investigated evolutionary patterns in traits related to epiphytic growth at the species and community level using a phylogeny including representatives of multiple epiphytic fern radiations on the island of Moorea, French Polynesia. 
\item We observed phylogenetic conservatism in multiple traits related to epiphytic growth in fern sporophytes and gametophytes. Epiphytes had significantly shorter stipes, smaller rhizomes, and a higher frequency of non-cordate gametophytes while accounting for phylogeny. Epiphytic communities were clustered both phylogenetically and functionally, consistent with the hypothesis that the canopy acts as a filter due to its low humidity. 
\item Our results provide an important phylogenetic perspective on the role of functional traits in ferns, and pave the way for more detailed studies examining how niche space is filled within terrestrial and epiphytic habitats.
\end{itemize}

@@

The sporophyte traits that differed most strongly between epiphytic and terrestrial taxa were related to plant size, including stipe and frond length, frond width, and rhizome diameter. Stipe length in particular differed strongly between epiphytes and terrestrial species (\fig\ref{fig:CWM}B, \tab\ref{table:PIC_sporo}). Stipe length may be a critical trait controlling whole leaf hydraulic conductance; since ferns have resistive stipes in general, species growing in water-limited habitats are expected to decrease stipe length to minimize resistance \citep{WatkinsJr.2010}. Our results support previous non-phylogenetically controlled studies that found a prevalence of short stipes in epiphytic ferns \citep{WatkinsJr.2010,Creese2011}. However, given that many of these size-related traits scale together in ferns \citep{Arcand2008,Creese2011}, careful experimental manipulations are needed to distinguish the role (or lack thereof) of each in epiphytic growth.

Surprisingly, SLA was only slightly lower in epiphytes relative to terrestrial species (\tab\ref{table:PIC_sporo}, \fig\ref{fig:CWM}B). SLA is an important part of the leaf economic spectrum, which posits that leaf traits including SLA, N and C content, and leaf life span vary along well-defined axes such that plants generally fall into one of two categories: fast-growing species with high SLA, low nutrient content, and short life spans vs. slow-growing species with low SLA, high nutrient content, and long life spans \citep{Wright2004a}. We hypothesized that epiphytes would have lower SLA, given their more extreme growing conditions. Indeed, the species with the lowest SLA values were epiphytes (\eg, \textit{Ctenopterella blechnoides} (Grev.) Parris, \SI{\Sexpr{as.numeric(get_result(filter(sla.grand.mean.results, species=="Ctenopterella_blechnoides"), "mean", signif=TRUE))}} {\square\meter\per\kilo\gram}; \textit{Selliguea plantaginea} Brack., \SI{\Sexpr{as.numeric(get_result(filter(sla.grand.mean.results, species=="Selliguea_plantaginea"), "mean", signif=TRUE))}} {\square\meter\per\kilo\gram}), but this relationship was not significant across the dataset (\tab\ref{table:PIC_sporo}, \fig\ref{fig:CWM}). The tradeoffs posited in the leaf economics spectrum have been most thoroughly documented in global studies including the widest possible range of taxa and environments, and do not always hold up in local studies \citep{Petter2016}. For example, \citet{Zhu2016} did not find a significant relationship between SLA and leaf life span in a survey of 16 ferns across a disturbance (light) gradient in China. It is possible that other small-scale effects (\eg, microclimate) obscure this pattern in local studies such as ours and that of \citet{Zhu2016}. This is supported by the strong, parallel relationship in CWM values of SLA with elevation in both epiphytes and terrestrial species (\fig\ref{fig:CWM}A).

@@

% Table S5 
\begin{table}[ht]
\caption{Trait values and growth habit of ferns from Moorea, French Polynesia. For continuous trait values, means $\pm$ \stdev shown. For growth habit, E = epiphytic, T = terrestrial. For frond dissection, 1 = simple, 2 = pinnatifid or pinnatisect, 3 = 1-pinnate, 4 = 1- pinnate-pinnatifid, 5 = 2-pinnate, 6 = 2-pinnate-pinnatifid, 7 = 3- pinnate, 8 = 3-pinnate-pinnatifid, 9 = \textgreater 3-pinnate-pinnatifid, 10 = binpinnatifid or tripinnatifid.}
\label{supp_table:traitvals}
\end{table}

### 2018-03-23 supporting_information.Rnw
Add Pagel's correlated evo test to manuscript
Also, change font to TNR in MS and SI
Fixes to mean_trait_values
Move sourcing of mean_trait_values to top of analyses list, add gameto_source to mean_trait_values, fix caption in supporting information Table S5

print(xtable(PCA.results,
             label="supp_table:PCA",
             caption="PC loadings for principal components analyses (PCA) of traits related to epiphytic growth in ferns from Moorea, French Polynesia. For each PC, values of the three traits with the greatest loadings (absolute values) shown in bold."),
      sanitize.text.function = function (x) {x},
      sanitize.colnames.function = custom_cols,
      sanitize.rownames.function = custom_rows,
      include.rownames=TRUE,
      caption.placement="top")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%% References %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

print(xtable(PCA.results,
             label="supp_table:PCA",
             caption="PC loadings for principal components analyses (PCA) of traits related to epiphytic growth in ferns from Moorea, French Polynesia. For each PC, values of the three traits with the greatest loadings (absolute values) shown in bold."),
      sanitize.text.function = function (x) {x},
      sanitize.colnames.function = custom_cols,
      sanitize.rownames.function = custom_rows,
      include.rownames=TRUE,
      caption.placement="top")
@

% Table S5 
\begin{table}[ht]
\caption{Trait values and growth habit of ferns from Moorea, French Polynesia. For continuous trait values, means $\pm$ \stdev shown. Growth habit was coded as a binary character (terrestrial or epiphytic). In case of intermediate growth habits, connection to the soil was used as the criterion for selecting the character state; \eg, epipetric plants were treated as epiphytic, and hemi-epiphytes and climbing plants were treated as terrestrial. For frond dissection, 1 = simple, 2 = pinnatifid or pinnatisect, 3 = 1-pinnate, 4 = 1- pinnate-pinnatifid, 5 = 2-pinnate, 6 = 2-pinnate-pinnatifid, 7 = 3-pinnate, 8 = 3-pinnate-pinnatifid, 9 = \textgreater 3-pinnate-pinnatifid, 10 = binpinnatifid or tripinnatifid. Morphotype and presence/absence of gemmae, glands, and hairs are traits of gametophytes; other traits are from sporophytes. For sporophyte data sources, "M" indicates measurements made on specimens collected for this study; 1 = \cite{NationalMuseumofNatureandScience2008}, 2= \cite{Copeland1932}, 3= \cite{Palmer2003}, 4 = \cite{RoyalBotanicGardensandDomainTrust}, 5 = \cite{Brownsey1987}}. For gametophyte data sources, "L" indicates observations of gametophytes cultured from spores in the lab; "F" indicates observations of gametophytes in the field and identified to species using DNA barcodes \citep{Nitta2017}; "T" indicates characters scored according to taxonomy (genus or family); 6 = \cite{Nayar1971}; 7 = \cite{Lloyd1980}; 8 = \cite{Zhang2008}; 9 = \cite{Bierhorst1967}; 10 = \cite{Martin2006}; 11 = \cite{Tigerschiold1989}; 12 = \cite{Tigerschiold1990}; 13 = \cite{Atkinson1975}.
\label{supp_table:traitvals}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%% References %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
